---
title: Mutation Accumulation Experiments
output: 
  github_document:
    toc: true
---



[![DOI](https://zenodo.org/badge/923498484.svg)](https://doi.org/10.5281/zenodo.14847205)



<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

This is a package to analyse data generated from Mutation Accumulation Experiments. 


# Installation

Clone the repository on your computer using

```
git clone git@github.com:konkam/MutAccExperiments.git
```

or 

```
git clone https://github.com/konkam/MutAccExperiments.git

```

You should then install the present package on your computer, using a command such a the following, from inside the package folder:


```
Rscript -e "devtools::install_local('.')"
```

Alternatively, you may open the folder using the Rstudio editor and press `Ctrl + Alt + B`.


- Dependencies should be:
    - R package devtools
    - JAGS + the R package RJags (On ubuntu, jags is on the canonical repository, available by `apt install jags`)
    - R package tidyverse
    - R package coda

# Data format


```{r}
library(MutAccExperiments)
library(tidyverse)
```

```{r}
data(example_input_data)

input_data_onestrain <- example_input_data %>%
  filter(strain == first(strain))
```

The minimum data required to run the analysis is a data frame with the following columns:

```{r}
minimal_input_data_onestrain <- example_input_data %>%
  filter(strain == first(strain)) %>%
  select(mutation_id, m, n, t)
minimal_input_data_onestrain
```

```{r}
input_data_onestrain %>%
  check_input_format_GCM()
```


# Fitting the GCM model

The GCM model is defined as follows:

$$
\begin{align}
m_i &\sim \text{Poisson}(n_i \mu_i t_i) \quad \forall i = 1, \ldots, I\\
\log_{10}(\mu_i) &\sim \mathcal{N}(\nu, \sigma) \\
\mu_0 &\sim \mathcal{N}(-8, 3) \\
\sigma &\sim \text{t}^+(0, 3, 5)\\
\end{align}
$$

where $m_i$ is the number of mutations observed of the $i$-th type, $n_i$ is the number of possible mutation sites of type $i$ in the genome, $t_i$ is the number of
elapsed generations and $\mu_i$ is the mutation rate for mutations of type $i$. $n_i, t_i$ are introduced so that $\mu_i$ can be interpreted as a mutation rate per site and
per generation, which will be comparable over mutation types. $\nu$ is the mean of the $\log10$ mutation rate, and $\sigma$ is the standard deviation of the $\log10$ mutation rate.
$\text{t}^+$ denotes the t distribution truncated to positive values.

This is a hierarchical structure, where $\nu$ represents the average tendency of the $\log10$ mutation rates (over mutation
types), and $\sigma$ represents the variability of the $\log10$ mutation rates between mutation types. These two values caracterise the average tendency and the dispersion of the mutation
rates.

```{r}
fit_GCM_model <- input_data_onestrain %>%
  EstimateMusGCM_onestrain()
```

## Accessing posterior samples

```{r}
extract_posterior_samples(fit_GCM_model, type = "mu")
```

```{r}
extract_posterior_samples(fit_GCM_model, type = "hyperparameters")
```

```{r}
extract_posterior_samples(fit_GCM_model, type = "predictive")
```

```{r}
extract_posterior_samples(fit_GCM_model, type = "prior")
```

Note that the m_pred variables are samples from the posterior predictive distribution evaluated at the observed data points.

The variables with a suffix "_prior" are samples from the prior distribution.

You can also access the posterior samples at once:

```{r}
extract_posterior_samples(fit_GCM_model)
```

## Various MCMC and model checks

```{r}
fit_GCM_model %>%
  traceplot()
```

```{r}
fit_GCM_model %>%
  summary()
```
```{r}
fit_GCM_model %>%
  plot_prior_posterior()
```

```{r}
posterior_predictive_one_strain(fit_GCM_model)
```
```{r}
plot_posterior_predictive_onestrain(fit_GCM_model)
```


# Fitting the model with MMR saturation

Compared to the previous unconstrained model, the MMR saturation model assumes that the mutation rates of the different
strains are related, i.e. constrained by each other. Four strains are considered, wild-type (wt), MMR-deactivated (MMR-), proofreading-deficient (proof-) and MMR-deactivated and proofreading-deficient (MMR-proof-).
The model includes an explicit parameter for the
saturation mechanism, which would be equal to 0 in the absence of saturation. Therefore, this
parameter quantifies both the evidence for the existence of a saturation mechanism and its
magnitude.

n practice, for each mutation type we assume that there is a baseline error rate $\gamma_i$. These mutations may
be corrected by the Proofreading mechanism with a probability $1-q_{\text{proofreading},i}$, and further
corrected by the MMR mechanism with a probability $1-q_{\text{MMR},i}$. Therefore, in the wild-type strain,
the number of observed mutation is impacted by the error rate and the two repair mechanisms. This
error rate is observed directly for the mutant MMR-proof- for which both MMR and Proofreading are inactivated.
The strain MMR- has the baseline error rate moderated by the Proofreading mechanism, the strain
C* has the baseline error moderated by the MMR mechanism which can saturate. Saturation is
modelled as follows: a proportion $1-\theta$ of mutations are corrected by the MMR mechanism with
probability  $1-q_{\text{MMR},i}$, until saturation may occur. At this point, the remaining proportion of
mutations remain uncorrected. As mentioned above, if $\theta=0$, there is no saturation, and an estimate
of $\theta$ significatively different from 0 provides evidence for saturation. This results in the following
system of equations for the rate of mutation of type $i$ per site
per generation in the four strains:

$$
\begin{align}
\mu_{\text{MMR-proof-},i} &= \gamma_i \\
\mu_{\text{MMR-},i} &= \gamma_i q_{\text{proofreading},i} \\
\mu_{\text{proof-},i} &= \gamma_i \left(\theta + (1-\theta) q_{\text{MMR},i} \right) \\
\mu_{\text{wt},i} &= \gamma_i q_{\text{proofreading},i}q_{\text{MMR},i} \\
\end{align}
$$

We choose the following generative model:

$$
\begin{align}
m_{s, i} &\sim \text{Poisson}(n_{s, i} \mu_{s, i} t_{s, i}) \qquad \text{for } i = 1, \ldots, I, s = \text{wt}, \text{MMR-}, \text{proof-}, \text{MMR-proof-}\\
\log10 \mu_{s, i} &\sim \mathcal{N}(\nu_{s}, \sigma_{s})  \qquad \text{for } i = 1, \ldots, I, s = \text{wt}, \text{MMR-}, \text{MMR-proof-} \\
\nu_s &\sim \mathcal{N}(-8, 3) \qquad \text{for } s = \text{wt}, \text{MMR-}, \text{MMR-proof-}\\
\sigma_s &\sim \text{t}^+(0, 3, 5) \qquad \text{for } s = \text{wt}, \text{MMR-}, \text{MMR-proof-}\\
\theta &\sim \text{Beta}(1, 1)\\
\end{align}
$$


```{r}
minimal_input_data <- example_input_data %>%
  select(strain, mutation_id, m, n, t)

minimal_input_data
```

```{r}
fit_MMRsaturation_model <- EstimateMusMMRsaturation(example_input_data)
```

## Accessing posterior samples

```{r}
extract_posterior_samples(fit_MMRsaturation_model, type = "mu")
```

```{r}
extract_posterior_samples(fit_MMRsaturation_model, type = "hyperparameters")
```

```{r}
extract_posterior_samples(fit_MMRsaturation_model, type = "predictive")
```

Note that the m_pred variables are samples from the posterior predictive distribution evaluated at the observed data points.

The variables with a suffix "_prior" are samples from the prior distribution.

You can also access the posterior samples at once:

```{r}
extract_posterior_samples(fit_MMRsaturation_model)
```

## Various MCMC and model checks

```{r}
fit_MMRsaturation_model %>%
  traceplot()
```

```{r}
fit_MMRsaturation_model %>%
  summary()
```

```{r}
fit_MMRsaturation_model %>%
  plot_prior_posterior()
```

```{r}
fit_MMRsaturation_model %>%
  posterior_predictive()
```


```{r}
fit_MMRsaturation_model %>%
  plot_posterior_predictive()
```

