---
title: Mutation Accumulation Experiments
output: 
  github_document:
    toc: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

This is a package to analyse data generated from Mutation Accumulation Experiments. 


# Installation

Clone the repository on your computer using

```
git clone git@forgemia.inra.fr:konkam/MutAccExperiments.git
```

or 

```
git clone https://forgemia.inra.fr/konkam/MutAccExperiments.git

```

You should then install the present package on your computer, using a command such a the following, from inside the package folder:


```
Rscript -e "devtools::install_local('.')"
```

Alternatively, you may open the folder using the Rstudio editor and press `Ctrl + Alt + B`.


- Dependencies should be:
    - R package devtools
    - JAGS + the R package RJags (On ubuntu, jags is on the canonical repository, available by apt install jags)
    - R package tidyverse
    - R package coda
    - A functioning latex installation
    - cmake

# Data format


```{r}
library(MutAccExperiments)
library(tidyverse)
```

```{r}
input_data = read.delim(file = "data-raw/SAT_mcmc_chain_labmut_R3610MMR-3610_inputfile.csv", sep = " ") %>% 
  rename(mutation_id = context_id,
         m = m_sc, 
         n = n_c,
         t = t_s, 
         mutation_label = labmut) %>% 
  mutate(strain = gsub("MMR-", "MMRminus", strain))  %>%
  mutate(strain = gsub("WT3610", "wt", strain)) %>% 
  mutate(strain = gsub("polC_mutL", "MMRproofminus", strain)) %>% 
  mutate(strain = gsub("polC", "proofminus", strain))

input_data_onestrain = input_data %>% 
  filter(strain == first(strain)) 
```

The minimum data required to run the analysis is a data frame with the following columns:

```{r}
minimal_input_data_onestrain = input_data %>% 
  filter(strain == first(strain)) %>% 
  select(mutation_id, m, n, t) 
minimal_input_data_onestrain
```

```{r}
input_data_onestrain %>% 
  check_input_format_GCM()
```


# Fitting the GCM model

```{r}
fit_GCM_model = input_data_onestrain %>% 
  EstimateMusGCM_onestrain()
```

## Accessing posterior samples

```{r}
extract_posterior_samples(fit_GCM_model, type = "mu")
```

```{r}
extract_posterior_samples(fit_GCM_model, type = "hyperparameters")
```

```{r}
extract_posterior_samples(fit_GCM_model, type = "predictive")
```

```{r}
extract_posterior_samples(fit_GCM_model, type = "prior")
```

Note that the m_pred variables are samples from the posterior predictive distribution evaluated at the observed data points.

The variables with a suffix "_prior" are samples from the prior distribution.

You can also access the posterior samples at once:

```{r}
extract_posterior_samples(fit_GCM_model)
```

## Various MCMC and model checks

```{r}
fit_GCM_model %>% 
  traceplot
```

```{r}
fit_GCM_model %>% 
  summary
```
```{r}
fit_GCM_model %>% 
  plot_prior_posterior
```

```{r}
posterior_predictive_one_strain(fit_GCM_model)
```
```{r}
plot_posterior_predictive_onestrain(fit_GCM_model)
```


# Fitting the model with MMR saturation

```{r}
minimal_input_data = input_data %>% 
  select(strain, mutation_id, m, n, t) 

minimal_input_data

```

```{r}
fit_MMRsaturation_model = EstimateMusMMRsaturation(input_data)
```

## Accessing posterior samples

```{r}
extract_posterior_samples(fit_MMRsaturation_model, type = "mu")
```

```{r}
extract_posterior_samples(fit_MMRsaturation_model, type = "hyperparameters")
```

```{r}
extract_posterior_samples(fit_MMRsaturation_model, type = "predictive")
```

Note that the m_pred variables are samples from the posterior predictive distribution evaluated at the observed data points.

The variables with a suffix "_prior" are samples from the prior distribution.

You can also access the posterior samples at once:

```{r}
extract_posterior_samples(fit_MMRsaturation_model)
```

## Various MCMC and model checks

```{r}
fit_MMRsaturation_model %>% 
  traceplot
```

```{r}
fit_MMRsaturation_model %>% 
  summary
```
```{r}
fit_MMRsaturation_model %>% 
  plot_prior_posterior
```
```{r}
fit_MMRsaturation_model %>% 
  posterior_predictive
```


```{r}
fit_MMRsaturation_model %>% 
  plot_posterior_predictive
```

